\propertitle

### 2019-11-16 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-16T19:24:32+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{P_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\pxxx}{P}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycmm}{C}
\newcommand*{\ycm}{\mathte{\ycmm}}
\newcommand*{\ycs}{\bm{\sigma}}
\newcommand*{\yL}{L}
\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\yds}{\delta}

@@

A natural procedure to compare the informational sufficiency of two
different sets of moments $M'\defd\set{1,\dotsc,m'}$ and
$M''\defd \set{1,\dotsc,m''}$ is as follows:
\begin{enumerate*}[label=(\roman*)]
\item calculate the relative entropies of the measured frequencies $\yf$
  and the sample marginals of the maximum-entropy distributions built with
  the two corresponding sets of constraints;
\item multiply these relative entropies by the number of time bins of the
  recording;
\item take their difference.
\end{enumerate*}
In formulae,
\begin{multline}
  \label{eq:measure_suffic}
  \text{difference in informational sufficiency between $M'$ and $M''$} \defd{}\\
T  \sum_{\ya} \yff_{\ya} \ln\frac{\yff_{\ya}}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M')}
-
T  \sum_{\ya} \yff_{\ya} \ln\frac{\yff_{\ya}}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'')} \equiv{}\\
T  \sum_{\ya} \yff_{\ya} \ln\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'')}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M')}.
\end{multline}
Why is this a natural measure? Let's ask how much more probable is the
sufficiency of one set with respect to the other, conditional on our data
$\yf$:
\begin{equation}
  \label{eq:ratio_prob_suffic}
  \pf(M'' \| \yf)/
  \pf(M' \| \yf).
\end{equation}
Now, the probability of observing activity $\ya$ in the sample at any time
bin is the sample marginal of the maximum-entropy distribution for the full
network, owing to the excheangeability assumption implicit in the
maximum-entropy method:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya_{t} \| M) = \sum_{\yA}\yG_{\ya_{t}\,\yA}\px(\yA \| M).
\end{equation}
The probability of observing one sequence $(\ya_{t})$ with frequencies
$\yf$ is therefore
\begin{equation}
  \label{eq:prob_frequencies}
  \prod_{t=1}^{T} \pxx(\ya_{t} \| M) \equiv
  \prod_{\ya=0}^{n} \pxx(\ya \| M)^{T\,\yff_{\ya}} \equiv
  \prod_{\ya=0}^{n}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{T\,\yff_{\ya}}.
\end{equation}
The probability of observing the frequencies $\yf$ is obtained multiplying this
by their multiplicity factor, the multinomial coefficient
\begin{equation}
  \label{eq:multiplicity_factor}
  \binom{T}{T\yf} \defd \frac{T!}{\prod_{\ya}(T\,\yff_{\ya})!}
  \approx \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}},
\end{equation}
the last expression coming from Stirling's approximation
\citep[Lemma~2.2]{csiszaretal2004b}.
If we assign equal pre-data probabilities to the two hypotheses $M'$ and $M''$,
each probability in the ratio~\eqref{eq:ratio_prob_suffic} then becomes, by
Bayes's theorem,
\begin{multline}
  \label{eq:prob_proportional_relentropy}
  \pf(M \| \yf) \propto
  \pf(\yf \| M) \times\text{\small const} \propto
  \binom{T}{T\yf}\,  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}
   \approx{}\\
  \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}}
  \times
  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}.
\end{multline}
The logarithm of the probability above is easily seen to be the number of
bins $T$ multiplied by relative entropy between the frequency distribution
$\yf$ and the sample marginal of the maximum-entropy distribution.

@@

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-17T00:27:31+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{P_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\pxxx}{P}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycmm}{C}
\newcommand*{\ycm}{\mathte{\ycmm}}
\newcommand*{\ycs}{\bm{\sigma}}
\newcommand*{\yL}{L}
%\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\ydi}{\varDelta}

@@

We measure the difference $\ydi(M'',M')$ in informational sufficiency
between a set of moments, say $M''\defd\set{1,\dotsc,m''}$, and another, say
$M'\defd \set{1,\dotsc,m'}$, as follows:
\begin{enumerate}[label=(\roman*)]
\item from each maximum-entropy distributions $\px(\yA \| M)$ for the full
  network, built from each set of constraints $M=M',M''$, calculate the
  marginal distribution for the sample:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M),
  \qquad M=M',M'';
\end{equation}
\item calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal, and multiply
  them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M)] \defd  T  \sum_{\ya} \yff_{\ya} \ln\frac{\yff_{\ya}}{
      \pxx(\ya \| M)},
    \qquad M=M',M'';    
  \end{equation}
\item take the difference: 
\begin{multline}
  \label{eq:measure_suffic}
  \ydi(M'',M') \defd
  T \,\sh[\yf; \pxx(\ya \| M')]
  -
  T \,\sh[\yf; \pxx(\ya \| M'')]
  \equiv{}\\
T  \sum_{\ya} \yff_{\ya} \ln\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'')}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M')}.
\end{multline}
\end{enumerate}
The measure $\ydi(M'',M')$ so defined is positive if $M''$ is \enquote{more
  informationally sufficient} than $M'$.

Why is this a natural measure? Because, if we assign equal pre-data
probabilities to the two hypotheses $M''$ and $M'$, then the exponential of
$\ydi(M'',M')$ is approximately equal to the ratio of their probabilities
conditional on the data $\yf$ -- their Bayes factor:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi(M'',M') \approx
  \ln\bigl[\pf(M'' \| \yf)/
  \pf(M' \| \yf)\bigr].
\end{equation}
So the exponential of $\ydi(M'',M')$ tells us how much more probable $M''$
is than $M'$, in view of the data $\yf$. We prove this in appendix**

@@

Let's ask how much more probable is the
sufficiency of one set with respect to the other, conditional on our data
$\yf$:
\begin{equation}
  \label{eq:ratio_prob_suffic}
  \pf(M'' \| \yf)/
  \pf(M' \| \yf).
\end{equation}
Now, the probability of observing activity $\ya$ in the sample at any time
bin is the sample marginal of the maximum-entropy distribution for the full
network, owing to the excheangeability assumption implicit in the
maximum-entropy method:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya_{t} \| M) = \sum_{\yA}\yG_{\ya_{t}\,\yA}\px(\yA \| M).
\end{equation}
The probability of observing one sequence $(\ya_{t})$ with frequencies
$\yf$ is therefore
\begin{equation}
  \label{eq:prob_frequencies}
  \prod_{t=1}^{T} \pxx(\ya_{t} \| M) \equiv
  \prod_{\ya=0}^{n} \pxx(\ya \| M)^{T\,\yff_{\ya}} \equiv
  \prod_{\ya=0}^{n}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{T\,\yff_{\ya}}.
\end{equation}
The probability of observing the frequencies $\yf$ is obtained multiplying this
by their multiplicity factor, the multinomial coefficient
\begin{equation}
  \label{eq:multiplicity_factor}
  \binom{T}{T\yf} \defd \frac{T!}{\prod_{\ya}(T\,\yff_{\ya})!}
  \approx \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}},
\end{equation}
the last expression coming from Stirling's approximation
\citep[Lemma~2.2]{csiszaretal2004b}.
If we assign equal pre-data probabilities to the two hypotheses $M'$ and $M''$,
each probability in the ratio~\eqref{eq:ratio_prob_suffic} then becomes, by
Bayes's theorem,
\begin{multline}
  \label{eq:prob_proportional_relentropy}
  \pf(M \| \yf) \propto
  \pf(\yf \| M) \times\text{\small const} \propto
  \binom{T}{T\yf}\,  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}
   \approx{}\\
  \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}}
  \times
  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}.
\end{multline}
The logarithm of the probability above is easily seen to be the number of
bins $T$ multiplied by relative entropy between the frequency distribution
$\yf$ and the sample marginal of the maximum-entropy distribution.
